---
title: "Research"
---

My main research focuses are in visual statistics, statistics philosophy, and computational statistics. This means I spend a lot of time thinking about what an observation/statistic is, and how we can best represent it visually. Accurate visual representations of our data are crucial for communicating complicated statistical concepts (both with ourselves and others), so good visualisation tools are a necessity for the future of statistics. 

Visualisations of random variables are, in of themselves, random variables, with their own statistical properties. Ensuring that these visualisations have *good* statistical properties is the key focus of my work. 


# Software
Usually any work I do will also have a software output to make implementing the theories easier. This is that software. 

```{r}
#| echo: false
#| message: false
#| eval: false

# This code is basically fully copied from Di's website repo. Thank you Di

library(tidyverse)
library(tsibble)
library(lubridate)
library(pkgsearch)
# remotes::install_github("robjhyndman/pkgmeta")
library(pkgmeta)

# Functions for getting downloads
find_cran_packages <- function(name) {
  pkgsearch::ps(name, size = 100) %>%
    filter(purrr::map_lgl(
      package_data, ~ grepl(name, .x$Author, fixed = TRUE)
    )) %>%
    select(package) %>%
    pull(package)
}

# Monthly download counts from packages in <x>
cran_downloads <- function(x) {
  # Compute monthly download counts
  down <- cranlogs::cran_downloads(x, from = "2000-01-01") %>%
    as_tibble() %>%
    mutate(month = tsibble::yearmonth(date)) %>%
    group_by(month) %>%
    summarise(count = sum(count), package = x)
  # Strip out initial zeros
  first_nonzero <- down %>%
    filter(count > 0) %>%
    head(1)
  if (NROW(first_nonzero) > 0) {
    filter(down, month >= first_nonzero$month)
  } else {
    first_nonzero
  }
}

packages <- find_cran_packages("Harriet Mason")

pkg_data <- pkgmeta:::get_meta_cran(packages,
    include_downloads=TRUE) |>
    rename(released = first_download, last_update = date)

# add hex file
hex <- c("hex/cassowaryr.png", "hex/ggdibbler.png")
pkg_data <- cbind(pkg_data, hex)

# Save data
save(pkg_data, file = "software.rda")
```




```{r}
#| echo: false
#| message: false
library(readr)
library(dplyr)
library(glue)
library(stringr)
library(flextable)
library(magick)

load("software.rda")

pkg_tbl <- pkg_data |>
   arrange(desc(last_update)) |>
  select(package, last_update, released, version, title, downloads, url, hex) |>
  flextable() |>
    colformat_num(big.mark = ",", decimal.mark = ".") |>
    colformat_date(fmt_date = "%d/%m/%Y")
pkg_tbl <- compose(
  x = pkg_tbl,
  j = "package",
  value = as_paragraph(hyperlink_text(x=package, url=url,
     props = officer::fp_text(color = "#a96469")))) |>
  delete_columns(j = "url")

pkg_tbl <- colformat_image(x = pkg_tbl,
                           i=1:nrow(pkg_data),
                           j = "hex", width = 1, height = 1.156,
                           na_str = "")
pkg_tbl
```


# Talks

These are the links to the slides for (almost) all the presentations I have given on my research.

```{r talks}
#| echo: false
#| message: false

library(readr)
library(dplyr)
library(gt)
library(glue)
library(lubridate)

# data from cv
talks  <- tribble(
  ~ date, ~ title, ~ venue, ~ link,
  2025, "Visualising Uncertainty for Holistic Insights", "ASC", "https://harriet-mason.github.io/talk-asc2025/",
  2025, "Visualising Uncertainty with ggdibbler", "WOMBAT", "https://harriet-mason.github.io/talk-wombat-2025/",
  2025, "Visualising Uncertainty with ggdibbler", "useR!", "https://harriet-mason.github.io/talk_useR-2025/",
  2024, "The Noisy Work of Uncertainty Visualisation Research", "AMSI Winter School", NA,
  2023, "Teaching Computers to See Patterns in Scatterplots", "R Ladies, Melbourne", NA,
  2022, "Teaching Computers to See Patterns in Scatterplots", "useR!", "https://github.com/harriet-mason/UserTalk_2022",
  )

talks <- talks |>
  mutate(titlelink = ifelse(is.na(link), title, glue("[{title}]({link})")))
talks |>
  arrange(desc(date)) |>
  select(date, titlelink, venue) |>
  gt() |>
    gt::fmt_markdown(columns = titlelink) |>
    cols_label(titlelink = "title") |>
    gt::cols_width(
      date ~ px(100),
      titlelink ~ px(400),
      venue ~ px(250)) |>
  cols_align("left")
```

